Kernel k-Groups and Hartigan's Method

### 2018-06-16 energy_clustering/pami18/energy_clustering.tex
pami

\begin{algorithm}
\begin{algorithmic}[1]
    \INPUT $k$, $G$, $\mathcal{W}$, $Z \leftarrow Z_0$
    \OUTPUT $Z$ 
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            \hfill (see \eqref{eq:maxQ})
  \STATE $s \leftarrow (s_1,\dotsc,s_k)^\top$ 
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i)$
            \hfill (see \eqref{eq:Jell})
        \IF{ $j^\star \ne j$} 
            \STATE $Z_{ij} \leftarrow 0$ 
            \STATE $Z_{ij^\star} \leftarrow 1$
            \STATE $s_j \leftarrow s_j - \mathcal{W}_{ii}$ 
            \STATE $s_{j^\star} \leftarrow s_{j^\star} + \mathcal{W}_{ii}$
            \STATE $q_j \leftarrow q_j - 2Q_j(x_i)$ 
            \hfill (see \eqref{eq:costxij})
            \STATE $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
            \hfill (see \eqref{eq:costxij})
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{kmeans_algo}
Weighted version of kernel k-means algorithm
to find local solutions to the optimization problem 
\eqref{eq:qcqp3}.
}
\end{algorithm}

@@

@@

\begin{algorithm}
\begin{algorithmic}[1]
    \INPUT $k$, $G$, $\W$, $Z \leftarrow Z_0$
    \OUTPUT $Z$
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            \hfill (see \eqref{eq:maxQ})
  \STATE $s \leftarrow (s_1,\dotsc,s_k)^\top$ 
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
                \Delta Q^{j\to \ell}(x_i)$
            \hfill (see \eqref{eq:changeQ}) \label{stepmove}
        \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
            \STATE $Z_{ij} \leftarrow 0$
            \STATE $Z_{ij^\star} \leftarrow 1$
            \STATE $s_j \leftarrow s_j - \mathcal{W}_{ii}$ 
            \STATE $s_{j^\star} \leftarrow s_{j^\star} + \mathcal{W}_{ii}$
            \STATE $q_j \leftarrow q_j - 2Q_j(x_i) + (\W G \W)_{ii}$
                \hfill (see \eqref{eq:costxij})
            \STATE 
            $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)+ 
                (\W G\W)_{ii}$
                \hfill (see \eqref{eq:costxij})
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{algo}
Kernel k-groups algorithm, based on Hartigan's method, to find
local solutions to problem \eqref{eq:qcqp3}.
%Steps $6$, $12$ and $13$ are different
%compared to Algorithm~\ref{kmeans_algo}. 
}
\end{algorithm}

@@

%In the first two experiments 
%we use $\mathcal{E}^H$-clustering with the standard metric from energy
%statistics, i.e. \eqref{eq:rho_alpha} with \mbox{$\alpha=1$}. 
%We consider univariate
%normal and lognormal mixtures according to
%\begin{align}
%x &\stackrel{iid}{\sim} \tfrac{1}{2}
%\mathcal{N}(\mu_1,\sigma_1^2) + 
%\tfrac{1}{2} \mathcal{N}(\mu_2,\sigma_2^2) ,\label{eq:uni_normal} \\
%x & \stackrel{iid}{\sim} \tfrac{1}{2}
%\exp\left\{ \mathcal{N}(\mu_1,\sigma_1^2) \right\} + 
%\tfrac{1}{2} \exp \left\{ \mathcal{N}(\mu_2,\sigma_2^2)\right\} ,
%\label{eq:uni_lognormal} \\
%\mu_1 &= 1.5, \ \sigma_1=0.3, \ \mu_2=0, \ \sigma_2=1.5.
%\label{eq:uni_params}
%\end{align}
%We increase the sample size $n$ and show the clustering accuracy 
%\eqref{eq:accuracy}
%versus $n$ in Fig.~\ref{fig:1D}. 
%In both cases the optimal accuracy 
%obtained from Bayes classification error
%is $ \approx 0.88$, indicated by the dashed line.
%Note that $\mathcal{E}^H$-clustering 
%performs better than k-means for normally distributed data, 
%and worse than GMM,
%as expected. 
%However, for lognormal distributions k-means and GMM are only slightly
%better than chance while $\mathcal{E}^H$-clustering is still 
%accurate. The reason is that k-means and GMM models are inconsistent
%with this lognormal data. Energy statistics on the other hand
%is model-free. We also show 
%a density estimation after clustering $1000$ points from these
%mixtures.
%Note that only $\mathcal{E}^{H}$-clustering
%was able to distinguish between the two classes, while 
%k-means and GMM put most of the
%points in a single cluster and points on the tail of the second component
%of \eqref{eq:uni_lognormal} in the other cluster.
%
%
%\begin{figure}
%\begin{minipage}{0.49\linewidth}
%\begin{center}
%\includegraphics[width=\textwidth]{1D_normal.pdf}\\
%~~~~~\includegraphics[width=.9\textwidth]{normal_density.pdf}
%\end{center}
%\end{minipage}
%\begin{minipage}{0.49\linewidth}
%\begin{center}
%\includegraphics[width=\textwidth]{1D_lognormal.pdf}\\
%~~~~~\includegraphics[width=.9\textwidth]{lognormal_density.pdf}
%\end{center}
%\end{minipage}
%%\put(-140,5){\scriptsize{(a)}}
%%\put(-140,-78){\scriptsize{(c)}}
%%\put(-10,5){\scriptsize{(b)}}
%%\put(-10,-78){\scriptsize{(d)}}
%\caption{
%\label{fig:1D}
%\emph{Top Left:} Clustering data from \eqref{eq:uni_normal}.
%\emph{Top Right:} Clustering data from \eqref{eq:uni_lognormal}.
%The dashed lines show the optimal Bayes accuracy, which in both cases 
%are $\approx 0.88$.
%\emph{Bottom Left:} 
%Density estimation of each component in the mixture \eqref{eq:uni_normal} 
%after clustering $1000$ sampled points.
%\emph{Bottom Right:} 
% The same but for lognormal data \eqref{eq:uni_lognormal}.
%}
%\end{figure}

First, we analyze
how the algorithms degrade as the number of dimensions increase.
Consider data from the Gaussian mixture
\begin{equation}
\label{eq:gauss1}
\begin{split}
x  &\stackrel{iid}{\sim} 
\tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1) +
\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2), \quad
\Sigma_1=\Sigma_2 = I_D, \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top, \qquad
\mu_2 = 0.7 (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top.
\end{split}
\end{equation}
The Bayes error is fixed as $D$ increases giving an optimal accuracy
of $\approx 0.86$.
We sample $200$ points on each trial.
The results are shown in Fig.~\ref{fig:gauss1}.
We see that 
kernel k-groups and spectral clustering have close
performance, being superior to kernel k-means, k-means, and GMM.
The improvement is noticeable in 
higher dimensions.

\begin{figure}
\centering
\includegraphics[scale=0.5]{gauss1.pdf}
\caption{
\label{fig:gauss1}
High dimensional Gaussian mixture according to \eqref{eq:gauss1}.
We sample $200$ points total on each trial. The dashed line is
Bayes accuracy $\approx 0.86$. We use the metric $\rho_1$ in
\eqref{eq:rho_alpha}, which is standard in energy statistics.
}
\end{figure}

Still for a two-class Gaussian mixture as in \eqref{eq:gauss1}, we now choose
different numbers for the  diagonal covariance $\Sigma_2$.
We have $\Sigma_1=I_D$, $\mu_1=(0,\dotsc,0)^\top \in \mathbb{R}^D$,
$\mu_2=(1,\dotsc,1,0,\dotsc,0)^T \in \mathbb{R}^D$, 
with signal in the first $10$ dimensions, and
\begin{equation}
\label{eq:gauss2}
\begin{split}
\Sigma_2 &= \left( \begin{array}{c|c}
\widetilde{\Sigma}_{10} & 0 \\ \hline 
0 & I_{D-10} \end{array}\right), \\
\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,\\
&\hspace{3.6em}1.969, 4.035,   4.237,  2.813,  3.637).
\end{split}
\end{equation}
We simply chose $10$ numbers uniformly at random on the interval
$[1,5]$ and other choice would give analogous results.
Bayes  accuracy is fixed at $\approx 0.95$.
From Fig.~\ref{fig:gauss2} we see that 
all the methods are similarly accurate in low dimensions, but they quickly
degenerate as the number of dimensions increase, except 
kernel k-groups which is much more stable.

\begin{figure}
\centering
\includegraphics[scale=0.5]{gauss2.pdf}
\caption{
\label{fig:gauss2}
High dimensional Gaussian mixture according to \eqref{eq:gauss2}.
We sample $200$ points total on each trial. The dashed line is
Bayes accuracy $\approx 0.95$. We use $\rho_1$ in \eqref{eq:rho_alpha}
which standard metric in energy statistics.
}
\end{figure}

Now, consider
$x \stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1)+
\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2)$
with 
\begin{equation}
\label{eq:20gauss}
\begin{split}
2\Sigma_1 &= \Sigma_2 = I_{20} \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top , 
\quad \mu_2 = \tfrac{1}{2} 
(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top.
\end{split}
\end{equation}
Bayes accuracy is $\approx 0.90$. 
We increase the sample size $n \in [10, 400]$ and show
the accuracy versus $n$ in Fig.~\ref{fig:gauss_n}. We compare
kernel k-groups, with different metrics,
to k-means and GMM. We also use
the best metric in this example for spectral clustering.
We notice a superior performance of kernel k-groups 
compared to the other methods.

\begin{figure}
\centering
\includegraphics[scale=0.5]{gauss_n.pdf}
\caption{
\label{fig:gauss_n}
Gaussian mixture with parameters \eqref{eq:20gauss}. We increase
the number of sampled points in each trial. We use different
metrics; see \eqref{eq:rho_alpha}--\eqref{eq:rho_hat}. 
Here, kernel k-groups is more accurate than spectral clustering.
}
\end{figure}

To consider non-normal data, we sample from the lognormal mixture 
$x \stackrel{iid}{\sim} \tfrac{1}{2} 
e^{\mathcal{N}(\mu_1,\Sigma_1)}+
\tfrac{1}{2} e^{\mathcal{N}(\mu_2,\Sigma_2)}$ with the same
parameters as in \eqref{eq:20gauss}. 
The optimal Bayes accuracy is still $\approx 0.9$.
The results are shown in Fig.~\ref{fig:loggauss_n}.
We use exactly the same metrics as in the normal
mixture of Fig.~\ref{fig:gauss_n} to illustrate that the 
proposed method still performs accurately.

\begin{figure}
\centering
\includegraphics[scale=0.5]{loggauss_n.pdf}
\caption{
\label{fig:loggauss_n}
Same experiment as in Fig.~\ref{fig:gauss_n} but with a 
lognormal mixture with parameters \eqref{eq:20gauss}.
Again, kernel k-groups is more accurate than alternatives.
The figure suggests that neither of these methods are
consistent on this example since Bayes accuracy is $\approx 0.90$.
}
\end{figure}

Finally, we show a limitation of kernel k-groups, which is shared
between 
all the other methods except for GMM.
For highly unbalanced clusters, k-means, spectral
clustering, kernel k-means and kernel k-groups all degenerate
more quickly than GMM. This is illustrated in Figure~\ref{fig:gauss_unbal}
where we generate data according to
\begin{equation}
\label{eq:gauss3}
\begin{split}
x &\stackrel{iid} \sim  
\dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1)+
\dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1), \\
\mu_1 &= (0,0,0,0)^\top , \quad
\mu_2 = 1.5\times (1,1,0,0)^\top, \\
\Sigma_1 &= I_4, \quad
\Sigma_2 = \left( 
\begin{array}{c|c} 
\tfrac{1}{2} I_2 & 0  \\ \hline
0 & I_2 
\end{array}\right), \\
n_1 &= N - m, \quad  n_2 = N + m, \quad N=300.
\end{split}
\end{equation}
We then increase $m \in [0,240]$ making
the clusters progressively more unbalanced.
For highly unbalanced clusters, we see that GMM performs better than
the other methods, which have basically similar performance.
Based on this experiment, an interesting problem would be to
extend kernel k-groups to account for unbalanced clusters.

\begin{figure}
\centering
\includegraphics[scale=0.5]{gauss_unbal.pdf}
\caption{
\label{fig:gauss_unbal}
Comparison between clustering methods
on unbalanced clusters. The data is normally 
distributed as \eqref{eq:gauss3} where
we vary $m \in [0, 240]$. We use the standard metric $\rho_1$ (see
\eqref{eq:rho_alpha}) from energy statistics.
}
\end{figure}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Discussion}
\label{sec:conclusion}

We proposed a formulation to clustering based on a weighted
version of energy statistics, 
valid for arbitrary spaces of negative type.
Our mathematical formulation of energy clustering 
reduces to a QCQP in the associated RKHS, as demonstrated in 
Proposition~\ref{th:qcqp3}.
We showed that the optimization problem
is equivalent
to kernel k-means, once the kernel is fixed, and also
to several graph partitioning problems.
Energy statistics, however, fixes
a family of standard kernels in Euclidean space, and
more general kernels 
on spaces of negative type can also be obtained.

We extended Hartigan's method to kernel spaces and proposed
Algorithm~\ref{algo}, which we call kernel k-groups.
This method was compared to kernel k-means and spectral clustering, besides
k-means and GMM. Our numerical results show a superior performance of the
proposed method, specially in high dimensions.
We stress that kernel k-groups has the same complexity of kernel k-means.

Kernel k-groups suffers a limitation shared by kernel k-means and spectral
clustering which involves highly unbalanced
clusters. An interesting problem  that we leave open 
would is to extend the method to such 
situations, as well as a soft version.
Finally, kernel methods can benefit from sparsity and
fixed-rank approximations of the Gram matrix, and there is plenty
of room to make kernel k-groups more scalable.

@@

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
%\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

@@

>>>>

\begin{algorithm}[t]
\begin{algorithmic}[1]
    \INPUT $k$, $G$, $\mathcal{W}$, $Z \leftarrow Z_0$
    \OUTPUT $Z$ 
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            \hfill (see \eqref{eq:maxQ})
  \STATE $s \leftarrow (s_1,\dotsc,s_k)^\top$ 
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmin_{\ell=1,\dotsc,k} J^{(\ell)}(x_i)$
            \hfill (see \eqref{eq:Jell})
        \IF{ $j^\star \ne j$} 
            \STATE $Z_{ij} \leftarrow 0$ 
            \STATE $Z_{ij^\star} \leftarrow 1$
            \STATE $s_j \leftarrow s_j - \mathcal{W}_{ii}$ 
            \STATE $s_{j^\star} \leftarrow s_{j^\star} + \mathcal{W}_{ii}$
            \STATE $q_j \leftarrow q_j - 2Q_j(x_i)$ 
            \hfill (see \eqref{eq:costxij})
            \STATE $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)$
            \hfill (see \eqref{eq:costxij})
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{kmeans_algo}
Weighted version of kernel k-means algorithm
to find local solutions to the optimization problem 
\eqref{eq:qcqp3}.
}
\end{algorithm}

@@

\begin{algorithm}[t]
\begin{algorithmic}[1]
    \INPUT $k$, $G$, $\W$, $Z \leftarrow Z_0$
    \OUTPUT $Z$
  \STATE $q \leftarrow (Q_1, \dotsc, Q_k)^\top$ 
            \hfill (see \eqref{eq:maxQ})
  \STATE $s \leftarrow (s_1,\dotsc,s_k)^\top$ 
  \REPEAT
    \FOR{ $i=1,\dotsc,n$}
        \STATE let $j$ be such that $x_i \in \C_j$
        \STATE $j^\star \leftarrow \argmax_{\ell=1,\dotsc,k \, | \, \ell\ne j} 
                \Delta Q^{j\to \ell}(x_i)$
            \hfill (see \eqref{eq:changeQ}) \label{stepmove}
        \IF{ $\Delta Q^{j \to j^\star}(x_i) > 0$ }
            \STATE $Z_{ij} \leftarrow 0$
            \STATE $Z_{ij^\star} \leftarrow 1$
            \STATE $s_j \leftarrow s_j - \mathcal{W}_{ii}$ 
            \STATE $s_{j^\star} \leftarrow s_{j^\star} + \mathcal{W}_{ii}$
            \STATE $q_j \leftarrow q_j - 2Q_j(x_i) + (\W G \W)_{ii}$
                \hfill (see \eqref{eq:costxij})
            \STATE 
            $q_{j^\star} \leftarrow q_{j^\star} + 2Q_{j^\star}(x_i)+ 
                (\W G\W)_{ii}$
                \hfill (see \eqref{eq:costxij})
    %    \ELSE
    %        \STATE Do nothing;
        \ENDIF
    \ENDFOR
  \UNTIL{convergence}
\end{algorithmic}
\caption{\label{algo}
Kernel k-groups algorithm, based on Hartigan's method, to find
local solutions to problem \eqref{eq:qcqp3}.
%Steps $6$, $12$ and $13$ are different
%compared to Algorithm~\ref{kmeans_algo}. 
}
\end{algorithm}

@@

@@

\begin{figure*}
\centering
\begin{minipage}{.4\textwidth}
\includegraphics[width=1\textwidth]{gauss_1d.pdf}\\[-2em]
(a)
\end{minipage}
\begin{minipage}{.4\textwidth}
\includegraphics[width=1\textwidth]{loggauss_1d.pdf} \\[-2em]
(b)
\end{minipage}\\[1em]
\begin{minipage}{.37\textwidth}
\includegraphics[width=1\textwidth]{normal_density.pdf}\\[-2em]
(c)
\end{minipage}\hspace{2em}
\begin{minipage}{.37\textwidth}
\includegraphics[width=1\textwidth]{lognormal_density.pdf}\\[-2em]
(d)
\end{minipage}
\caption{
\label{fig:1dexperiment}
Clustering one-dimensional data for a two-class problem.
(a) $x \stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(0, 1.5) +
\tfrac{1}{2}\mathcal{N}(1.5,0.3)$. (b) $x \stackrel{iid}{\sim}
\tfrac{1}{2}e^{\mathcal{N}(0, 1.5)} + \tfrac{1}{2} e^{\mathcal{N}(1.5, 0.3)}$.
In both cases we plot the accuracy versus the total number of points.
(c) Kernel density estimation for a single trial with 2000 sampled
points from the
distribution in the case (a). (d) Kernel density estimation in a single
trial with 
2000 points sampled from the distribution in the case (b).
}
\end{figure*}

We first consider one-dimensional data for a two-class problem. We compare
kernel k-groups with k-means and GMM, as illustrated in 
Fig.~\ref{fig:1dexperiment}. We have two mixtures of normal and 
lognormal distributions. Notice that in the kernel density estimation plots
for lognormal distribution, only kernel k-groups was able to distinguish
between the two classes. The accuracy results for both density estimation
cases are in Table~\ref{tb:dens}. We remark that kernel k-groups in
this one-dimensional example performed the same as the deterministic
Algorithm~\ref{algo1d} introduced in the Appendix.

\begin{table}[h]
\centering
\caption{
\label{tb:dens} Accuracy results for the density
estimation of Fig.~\ref{fig:1dexperiment}d--e.
}
\begin{tabular}{@{}l|ll@{}}
       Method        & normal & lognormal \\ \midrule[.5pt]
        kmeans       & 0.773  & 0.502     \\
         GMM         & 0.881  & 0.513     \\
   kernel k-groups   & 0.802  & 0.847  
\end{tabular}
\end{table}

Next, we analyze
how the algorithms degrade as the number of dimensions increase.
Consider data from the Gaussian mixture
\begin{equation}
\label{eq:gauss1}
\begin{split}
x  &\stackrel{iid}{\sim} 
\tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1) +
\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2), \quad
\Sigma_1=\Sigma_2 = I_D, \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times D})^\top, \qquad
\mu_2 = 0.7 (\underbrace{1,\dots,1}_{\times 10},
\underbrace{0,\dots,0}_{\times (D-10)})^\top.
\end{split}
\end{equation}
The Bayes error is fixed as $D$ increases giving an optimal accuracy
of $\approx 0.86$.
We sample $200$ points on each trial. A scatter plot of the last two
dimensions that contains signal in $\mu_2$ is shown in
Fig.~\ref{fig:scatter}a.
The clustering results are shown in Fig.~\ref{fig:plots}a.
We see that 
kernel k-groups and spectral clustering have close
performance, being superior to kernel k-means, k-means, and GMM.
The improvement is noticeable in 
higher dimensions.

\begin{figure*}
\centering
\includegraphics[width=0.19\textwidth]{gauss_means_scatter.pdf}
\includegraphics[width=0.19\textwidth]{gauss_cov_scatter.pdf}
\includegraphics[width=0.19\textwidth]{gauss_diff_metrics.pdf}
\includegraphics[width=0.19\textwidth]{loggauss_diff_metrics.pdf}
\includegraphics[width=0.19\textwidth]{unbalanced_scatter.pdf}
\put(-500,-8){(a)}
\put(-398,-8){(b)}
\put(-298,-8){(c)}
\put(-195,-8){(d)}
\put(-95,-8){(e)}
\caption{
\label{fig:scatter}
Scatter plot of the last two dimensions where $\mu_2$ has signal.
Each plot has 200 points total.
(a) Data distributed as in \eqref{eq:gauss1}.
(b) Data distributed as in \eqref{eq:gauss2}.
(c) Data distributed as in \eqref{eq:20gauss}.
(d) Parameters as in \eqref{eq:20gauss} but for lognormal mixture.
(e) Data from \eqref{eq:gauss3} with $N = 300$ and $m=200$.
}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.33\textwidth]{gauss1.pdf}
\includegraphics[width=0.33\textwidth]{gauss2.pdf}
\includegraphics[width=0.33\textwidth]{gauss_n.pdf}
\includegraphics[width=0.33\textwidth]{loggauss_n.pdf}
\includegraphics[width=0.315\textwidth]{gauss_unbal.pdf}
\put(-400,130){(a)}
\put(-230,130){(b)}
\put(-50,130){(c)}
\put(-320,-8){(d)}
\put(-140,-8){(e)}
\caption{
\label{fig:plots}
Clustering results associated to the data illustrated in 
Fig.~\ref{fig:scatter}. For each experiment we
perform 100 Monte Carlo runs. Error bars indicate standard error.
(a)
High dimensional Gaussian mixture according to \eqref{eq:gauss1}.
The dashed line is
Bayes accuracy $\approx 0.86$. We use the metric $\rho_1$ in
\eqref{eq:rho_alpha}, which is standard in energy statistics.
(b)
High dimensional Gaussian mixture according to \eqref{eq:gauss2}.
Bayes accuracy $\approx 0.95$. We use $\rho_1$ in \eqref{eq:rho_alpha}.
(c)
Gaussian mixture with parameters \eqref{eq:20gauss}. We increase
the number of sampled points in each trial. We use different
metrics; see \eqref{eq:rho_alpha}--\eqref{eq:rho_hat}. 
Here, kernel k-groups is more accurate than spectral clustering.
(d)
Same experiment as in Fig.~\ref{fig:plots}c but with a 
lognormal mixture with parameters \eqref{eq:20gauss}.
Again, kernel k-groups is more accurate than alternatives.
The plot suggests that neither of these methods are
consistent on this example since Bayes accuracy is $\approx 0.90$.
(e)
Comparison between clustering methods
on unbalanced clusters. The data is normally 
distributed as \eqref{eq:gauss3} where
we vary $m \in [0, 240]$. We use the standard metric $\rho_1$ (see
\eqref{eq:rho_alpha}) from energy statistics.
}
\end{figure*}

Still for a two-class Gaussian mixture as in \eqref{eq:gauss1}, we now choose
different numbers for the  diagonal covariance $\Sigma_2$.
We have $\Sigma_1=I_D$, $\mu_1=(0,\dotsc,0)^\top \in \mathbb{R}^D$,
$\mu_2=(1,\dotsc,1,0,\dotsc,0)^T \in \mathbb{R}^D$, 
with signal in the first $10$ dimensions, and
\begin{equation}
\label{eq:gauss2}
\begin{split}
\Sigma_2 &= \left( \begin{array}{c|c}
\widetilde{\Sigma}_{10} & 0 \\ \hline 
0 & I_{D-10} \end{array}\right), \\
\widetilde{\Sigma}_{10} &= \diag(1.367,  3.175,  3.247,  4.403,  1.249,\\
&\hspace{3.6em}1.969, 4.035,   4.237,  2.813,  3.637).
\end{split}
\end{equation}
We simply chose $10$ numbers uniformly at random on the interval
$[1,5]$ and other choice would give analogous results.
Bayes  accuracy is fixed at $\approx 0.95$. In Fig~\ref{fig:scatter}b
we show a scatter plot of the 9th and 10th dimension.
From Fig.~\ref{fig:plots}b we see that 
all the methods are similarly accurate in low dimensions, but they quickly
degenerate as the number of dimensions increase, except 
kernel k-groups which is much more stable.

Now, consider
$x \stackrel{iid}{\sim} \tfrac{1}{2} \mathcal{N}(\mu_1,\Sigma_1)+
\tfrac{1}{2} \mathcal{N}(\mu_2,\Sigma_2)$
with 
\begin{equation}
\label{eq:20gauss}
\begin{split}
2\Sigma_1 &= \Sigma_2 = I_{20} \\
\mu_1 &= (\underbrace{0,\dotsc,0}_{\times 20})^\top , 
\quad \mu_2 = \tfrac{1}{2} 
(\underbrace{1,\dotsc,1}_{5},\underbrace{0,\dotsc,0}_{15})^\top.
\end{split}
\end{equation}
Bayes accuracy is $\approx 0.90$.  A scatter plot of the 4th and 5th
dimensions is shown in Fig.~\ref{fig:scatter}c.
We increase the sample size $n \in [10, 400]$ and show
the accuracy versus $n$ in Fig.~\ref{fig:plots}c. We compare
kernel k-groups, with different metrics,
to k-means and GMM. We also use
the best metric in this example for spectral clustering.
We notice a superior performance of kernel k-groups 
compared to the other methods.

To consider non-normal data, we sample from the lognormal mixture 
$x \stackrel{iid}{\sim} \tfrac{1}{2} 
e^{\mathcal{N}(\mu_1,\Sigma_1)}+
\tfrac{1}{2} e^{\mathcal{N}(\mu_2,\Sigma_2)}$ with the same
parameters as in \eqref{eq:20gauss}. 
The optimal Bayes accuracy is still $\approx 0.9$.
A scatter plot is in Fig.~\ref{fig:scatter}d and the
results are shown in Fig.~\ref{fig:plots}d.
We use exactly the same metrics as in the normal
mixture of Fig.~\ref{fig:plots}c to illustrate that the 
proposed method still performs accurately.

Finally, we show a limitation of kernel k-groups, which is shared
between 
all the other methods except for GMM.
For highly unbalanced clusters, k-means, spectral
clustering, kernel k-means and kernel k-groups all degenerate
more quickly than GMM. A scatter plot of the first two dimensions
is shown in Fig.~\ref{fig:scatter}e and the clustering results are
in Fig.~\ref{fig:plots}e,
where we generate data according to
\begin{equation}
\label{eq:gauss3}
\begin{split}
x &\stackrel{iid} \sim  
\dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1)+
\dfrac{n_1}{2N} \mathcal{N}(\mu_1,\Sigma_1), \\
\mu_1 &= (0,0,0,0)^\top , \quad
\mu_2 = 1.5\times (1,1,0,0)^\top, \\
\Sigma_1 &= I_4, \quad
\Sigma_2 = \left( 
\begin{array}{c|c} 
\tfrac{1}{2} I_2 & 0  \\ \hline
0 & I_2 
\end{array}\right), \\
n_1 &= N - m, \quad  n_2 = N + m, \quad N=300.
\end{split}
\end{equation}
We then increase $m \in [0,240]$ making
the clusters progressively more unbalanced.
For highly unbalanced clusters, we see that GMM performs better than
the other methods, which have basically similar performance.
Based on this experiment, an interesting problem would be to
extend kernel k-groups to account for unbalanced clusters.

In Fig.~\ref{fig:2dscatter} we show examples of two-dimensional
datasets whose clustering results are shown in Table~\ref{tb:cigar_circle}.
For kernel k-means and kernel k-groups we initialize at random.
We see that both methods perform closely, with higher accuracy
than the other ones.

\begin{figure}
\includegraphics[width=.24\textwidth]{2cigars.pdf}
\includegraphics[width=.24\textwidth]{2circles.pdf}
\put(-240,-8){(a)}
\put(-110,-8){(b)}
\caption{
\label{fig:2dscatter}
(a) Data distributed as 
$x \stackrel{iid}{\sim}
\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)+
\tfrac{1}{2}\mathcal{N}(\mu_1,\Sigma_1)$, 
$\mu_1= (0,0)^\top$, 
$\mu_2 = (6.5, 0)^\top$ and 
$\Sigma_1 = \Sigma_2 = \diag(1, 20)$. We sample $800$ points.
(b) Concentric circles with radius $r_1 = 1$ and $r_2 = 3$, with 
noise $0.2 \cdot \mathcal{N}(0, I_2)$. We sample 
800 points with probability $1/2$ for each class.
} 
\end{figure}

\begin{table}
\caption{
\label{tb:cigar_circle}
Clustering the data from Fig.~\ref{fig:2dscatter}.
For spcetral clustering, kernel k-means and kernel k-groups we use
the metric $\widetilde{\rho}_2$ (see \eqref{eq:rho_tilde}) for the data
in Fig.~\ref{fig:2dscatter}a, while $\widehat{\rho}_1$ (see \eqref{eq:rho_hat})
for the data in Fig.~\ref{fig:2dscatter}b. We have 800
points on each trial and 30 Monte Carlo runs for both datasets.
}
\centering
\begin{tabular}{@{}l|ll|ll@{}}
                   & \multicolumn{2}{c|}{Fig.~\ref{fig:2dscatter}a}  &
                     \multicolumn{2}{c}{Fig.~\ref{fig:2dscatter}b}  \\
       Method      & Accuracy & SEM     & Accuracy & SEM \\ \midrule[.5pt]
        kmeans       & 0.5330 & 0.0048  & 0.5210   & 0.0027 \\
         GMM         & 0.9292 & 0.0295  & 0.5330   & 0.0039 \\
 spectral clustering & 0.5775 & 0.0096  & 0.7253   & 0.0028 \\
    kernel k-means   & 0.9999 & 0.0002  & 0.9999   & $4.16\, 10^{-5}$ \\
   kernel k-groups   & 0.9987 & 0.0003  & 0.9999   & $4.16\, 10^{-5}$  
\end{tabular}
\end{table}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusion}
\label{sec:conclusion}

We proposed a formulation to clustering based on a weighted
version of energy statistics, 
valid for arbitrary spaces of negative type.
Our mathematical formulation of energy clustering 
reduces to a QCQP in the associated RKHS, as demonstrated in 
Proposition~\ref{th:qcqp3}.
We showed that the optimization problem
is equivalent
to kernel k-means, once the kernel is fixed, and also
to several graph partitioning problems.
%Energy statistics, however, fixes
%a family of standard kernels in Euclidean space, and
%more general kernels 
%on spaces of negative type can also be obtained.

We extended Hartigan's method to kernel spaces and proposed
Algorithm~\ref{algo}, which we called kernel k-groups.
This method was compared to kernel k-means and spectral clustering, besides
k-means and GMM. Our numerical results show a superior performance of the
proposed method, specially in high dimensions.
We stress that kernel k-groups has the same complexity of kernel k-means.

Kernel k-groups suffers a limitation shared by kernel k-means and spectral
clustering which involves highly unbalanced
clusters. An interesting problem  that we leave open 
is to extend the method to such 
situations.
Finally, kernel methods can benefit from sparsity and
fixed-rank approximations of the Gram matrix, and there is plenty
of room to make kernel k-groups more scalable.

@@

% if have a single appendix:
%\appendix[Proof of the Zonklar Equations]
% or
\appendix  % for no appendix heading
% do not use \section anymore after \appendix, only \section*
% is possibly needed

@@

\section*{Two-Class Problem in One Dimension}
\label{sec:twoclass}

Here we consider the simplest possible case which
is one-dimensional data and a two-class problem. We propose
an algorithm that does not depend on initialization. We used this simple
scheme to compare with kernel k-groups given
in algorithm \ref{algo}. Both algorithms have the same clustering
performance in the one-dimensional examples that we tested.
